Systemix Article: Combined MAG-SMAG Profiling

### 2024-05-10 R/paper.Rmd
update to manuscript version 1

```{r setup}
#knitr::opts_knit$set(root.dir = '/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/mags-all/analysis/')
knitr::opts_knit$set(root.dir = '/Users/moises/Downloads/mags-all/')

@@

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b)

#fig sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_s1.pdf", fig_s1, height = 8, width = 16)
#ggsave("fig_s1.png", fig_s1, height = 8, width = 16)
```

@@

```{r fig2}
# Merge plots using patchwork
fig_2 <- { { anova_boxplot_cu / anova_boxplot_fe +
    plot_layout(ncol = 1, widths = c(0.75)) } | anova_boxplot_s / kegg_bubbles_plot } +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_2.pdf", fig_2, height = 14, width = 18)
#ggsave("fig_2.png", fig_2, height = 14, width = 18)
```

## Section 3: SNP Calling

@@

```{r snps}
# Explore CDS annotations (from Bakta)
snp_refcds <- read.csv("5-SNPs/dndscv/buildref/cds_table.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Buildref testing for bacteria.. (1 bin)
#https://github.com/im3sanger/dndscv/issues/62
#buildref(cdsfile = "5-SNPs/dndscv/buildref/cds_node_25854.txt", genomefile = "5-SNPs/dndscv/buildref/node_25854.fa", outfile = "5-SNPs/dndscv/buildref/testref.rda")
#load("5-SNPs/dndscv/buildref/testref.rda")

# Buildref for all bins (very slow: ~2 hrs in building object)
#buildref(cdsfile = "5-SNPs/dndscv/buildref/cds_table.txt", genomefile = "5-SNPs/dndscv/buildref/S15_bins.fa", outfile = "5-SNPs/dndscv/buildref/S15_refcds.rda")

# Inspect refdb
load("5-SNPs/dndscv/buildref/S15_refcds.rda")
print(RefCDS[[1]])
rm(gr_genes, RefCDS)

# Get annotations from the VCF generated with SnpEff
#sampleID chr pos ref mut
snp_mutations <- read.csv("5-SNPs/dndscv/run/S15_mutations.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
rm(snp_mutations)

# Run driver discovery or genes under positive selection (PS) (runtime ~5 min)
#snp_dndsout <- dndscv(snp_mutations, refdb = "5-SNPs/dndscv/buildref/S15_refcds.rda", cv = NULL, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
#save(snp_dndsout, file = "5-SNPs/dndscv/run/S15_dndsout.rda")
#load("5-SNPs/dndscv/run/S15_dndsout.rda")

# Explore output object..
#https://github.com/im3sanger/dndscv/issues/6
#https://github.com/im3sanger/dndscv/issues/64
head(snp_dndsout$sel_cv) #dN/dS ratios at gene-level
head(snp_dndsout$annotmuts)
print(snp_dndsout$globaldnds) #driver globals
mean(snp_selcv$wmis_cv) #mis avg.

# Save dN/dS per-gene matrix
snp_selcv <- snp_dndsout$sel_cv
rm(snp_dndsout)

# Group global dN/dS ratios:
#https://github.com/im3sanger/dndscv/issues/63
#consider significant p-value -> pmis_cv < 0.05?
#snp_selcv %>% dplyr::filter(pmis_cv < 0.05) %>% nrow() #n=109523
snp_global_pos <- snp_selcv %>%
  dplyr::filter(wmis_cv > 1) %>%
  dplyr::summarise(genes = n())
snp_global_pos$label = "dN/dS > 1"
snp_global_neg <- snp_selcv %>%
  dplyr::filter(wmis_cv < 1 & wmis_cv != 0) %>%
  dplyr::summarise(genes = n())
snp_global_neg$label = "dN/dS < 1"
snp_global_neu <- snp_selcv %>%
  dplyr::filter(wmis_cv == 0) %>%
  dplyr::summarise(genes = n())
snp_global_neu$label = "dN/dS = 0"
snp_global_dnds <- rbind(snp_global_pos, snp_global_neg, snp_global_neu)
snp_global_dnds <- snp_global_dnds %>%
  dplyr::mutate(perc = round(genes * 100 / sum(snp_global_dnds$genes), 1)) %>% #add %
  dplyr::mutate(perc2 = paste0(genes, " (", perc, "%)")) #label to plot
rm(snp_global_pos, snp_global_neg, snp_global_neu) #clean memory

# Filter genes under PS:
#https://github.com/im3sanger/dndscv/issues/56
#wmis_cv: dN/dS for missense mutations
#pmis_cv: p-value < 0.05 is expected to occur by chance under neutrality (null hypothesis)
#p-values for multiple testing (q-values)
#qallsubs_cv = 144 genes; significant q-values (p-values adjusted for multiple testing)
snp_psgenes <- snp_selcv %>%
  dplyr::filter(wmis_cv > 1 & pmis_cv < 0.05) %>%
  dplyr::mutate(dndsr = n_mis / n_syn)
#snp_psgenes %>% dplyr::filter(dndsr == Inf) %>% nrow() #n=118

# Load contig to bin matrix
snp_bins <- read.csv("5-SNPs/dndscv/results/S15_DASTool_contig2bin.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Load gene name annotations (from genes.gff file)
snp_genids <- read.csv("5-SNPs/dndscv/results/gene_id_name.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

## FUNCTIONS ##
# Function that returns the contig where a gene was found
fx_dndsGene2Contig <- function(g) {
  c <- snp_refcds %>%
    dplyr::filter(gene.name == g) %>%
    dplyr::select(chr) %>%
    as.character()
  return(c)
}
# Function that returns the bin given a contig
fx_dndsContig2Bin <- function(c) {
  b <- snp_bins %>%
    dplyr::filter(contig == c) %>%
    dplyr::select(bin) %>%
    as.character()
  return(b)
}
# Function that gets gene names
fx_dndsGeneName <- function(g) {
  g <- snp_genids %>%
    dplyr::filter(gene.id == g) %>%
    dplyr::select(gene.name) %>%
    as.character()
  return(g)
}
# Function that returns a row with taxa given a bin
fx_dndsTaxaBin <- function(b) {
  t <- mag_samples %>%
    dplyr::mutate(bin = paste0(mag_name, ".", mag_id)) %>%
    dplyr::filter(bin == b) %>%
    dplyr::select(phylum, genus, species, completeness, contamination)
  return(t)
}
# Function that process the product tag of NS genes
fx_dndsNsPathway <- function(p) {
  if (grepl("copper", p, ignore.case = TRUE) == TRUE)
    return("Copper")
  if (grepl("nickel", p, ignore.case = TRUE) == TRUE)
    return("Nickel")
  if (grepl("sulfur", p, ignore.case = TRUE) == TRUE)
    return("Sulfur")
  if (grepl("mercury", p, ignore.case = TRUE) == TRUE)
    return("Mercury")
  if (grepl("nitrogen", p, ignore.case = TRUE) == TRUE)
    return("Nitrogen")
}
# Function that gets dndscv stats for NS genes
fx_dndsNsStats <- function(g) {
  s <- snp_selcv %>%
    dplyr::filter(gene_name == g) %>%
    dplyr::select(gene_name, wmis_cv, pmis_cv)
  return(s)
}
## END FUNCTIONS ##

@@

# Select PS and add taxonomic classification
snp_dnds <- snp_psgenes %>%
  #dplyr::filter(dndsr > 1) %>%
  dplyr::group_by(bin) %>%
  dplyr::summarise(num = n()) %>%
  dplyr::arrange(desc(num))

snp_taxabins <- tibble(phylum = character(), genus = character(), species = character(),
                       completeness = numeric(), contamination = numeric())
for (i in 1:nrow(snp_dnds)) {
  #print(sprintf("%i -> bin: %s", i, fx_dndsTaxaBin(snp_dnds$bin[i])))
  snp_taxabins <- rbind(snp_taxabins, fx_dndsTaxaBin(snp_dnds$bin[i]))
}
snp_dnds['phylum'] <- snp_taxabins$phylum
snp_dnds['genus'] <- snp_taxabins$genus
snp_dnds['species'] <- snp_taxabins$species
snp_dnds['completeness'] <- snp_taxabins$completeness
snp_dnds['contamination'] <- snp_taxabins$contamination
rm(snp_taxabins)

@@

# Subset to plot mutations -> synonymous/missense (non-syn.)/nonsense
snp_psgenes_sub <- snp_psgenes %>%
  dplyr::filter(wmis_cv > 5.5) %>%
  pivot_longer(cols = c(n_syn, n_mis, n_non), names_to = "mut", values_to = "n_mut")

# Barplot global dN/dS ratios
snp_plot_dndstot <- ggplot(snp_global_dnds, aes(x = genes, y = reorder(label, genes), fill = label)) +
  geom_col(position = "dodge2", fill = "#4c6194") +
  geom_text(aes(label = perc2), vjust = 0.5, hjust = -0.1, size = 2) +
  #scale_fill_manual(values = c("#c56c65", "#778ab3", "#529548")) +
  scale_x_continuous(limits = c(0, 125000)) +
  labs(x = "Number of genes", y = "") +
  theme_classic() +
  theme(legend.position = "none")
ggsave("5-SNPs/dndscv/results/plot/plot_dnds_global.pdf", snp_plot_dndstot, height = 3, width = 5)

# Plot global driver genes per taxa
snp_plot_psbins <- ggplot(snp_dnds, aes(x = num, y = reorder(bin, num), fill = bin)) +
  geom_bar(stat = "identity", fill = "#E31A1C") +
  geom_text(aes(label = num), vjust = 0.5, hjust = -0.5, size = 2.5) +
  labs(x = "Number of genes under PS", y = "Bin") +
  scale_x_continuous(limits = c(0, 120)) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  facet_grid(.~species)
ggsave("5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 8.5, width = 9.5)

# Lollipop plot of PS
#ggdotchart(snp_psgenes_sub, x = "name", y = "wmis_cv", color = "#E31A1C", rotate = TRUE, sorting = "descending", add = "segments", ggtheme = theme_pubr())
snp_plot_psgenes <- ggplot(snp_psgenes_sub,
                           aes(x = wmis_cv, y = reorder(name, wmis_cv)), size = wmis_cv) +
  geom_segment(aes(x = 0, xend = wmis_cv, y = name), color = "gray") +
  geom_point(alpha = 0.7, color = "#E31A1C") +
  geom_vline(xintercept = max(snp_psgenes$wmis_cv), linetype = 1, linewidth = 0.1) +
  labs(x = "PS strength", y = "Gene CDS") +
  theme_classic()
ggsave("5-SNPs/dndscv/results/plot/plot_dnds_psgenes.pdf", snp_plot_psgenes, height = 9, width = 7)

# Plot mutations/genes per type
snp_plot_muts <- ggplot(snp_psgenes_sub, aes(x = n_mut, y = reorder(name, wmis_cv), fill = mut)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("mis", "non", "syn"),
                    values = c("#5db4b2", "#b687b6", "#2374a4")) +
  labs(x = "Number of mutations", y = "") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.15),
        strip.placement = "outside",
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
#ggsave("5-SNPs/dndscv/results/plot/plot_dnds_psmuts.pdf", snp_plot_muts, height = 9, width = 7)

snp_plot_psgenes <- (snp_plot_psgenes | snp_plot_muts) +
  plot_layout(nrow = 1, ncol = 2, widths = c(0.55, 0.5))
ggsave("5-SNPs/dndscv/results/plot/plot_dnds_psgenmut.pdf", snp_plot_psgenes, height = 9, width = 9)

# Negative selected (NS) genes with dN/dS < 1 (selected metals)
snp_nsgenes <- read.csv("5-SNPs/dndscv/results/selected_cds_metals.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Parse product to pathway and add additional attributes (wmis_cv, pmis_cv)
#add dndsr, contig, bin**?
snp_nspath <- tibble(gene_name = character(), pathway = character())
snp_nsstat <- tibble(wmis_cv = numeric(), pmis_cv = numeric())
for (i in 1:nrow(snp_nsgenes)) {
  #work using dataframes, not lists
  snp_nspath <- snp_nspath %>%
    add_row(gene_name = snp_nsgenes$gene_name[i], pathway = fx_dndsNsPathway(snp_nsgenes$product[i]))
	snp_nsstat <- rbind(snp_nsstat, fx_dndsNsStats(snp_nsgenes$gene_name[i]))
}
snp_nsgenes <- left_join(snp_nsgenes, snp_nspath, by = c("gene_name"))
snp_nsgenes <- left_join(snp_nsgenes, snp_nsstat, by = c("gene_name"))
rm(snp_nspath, snp_nsstat)

# Remove NA cells and significance: ie. BAOFFNEK_01342 not found
snp_nsgenes <- snp_nsgenes %>%
  dplyr::filter(!is.na(pathway) & !(is.na(wmis_cv))) %>%
  dplyr::filter(wmis_cv != 0 & pmis_cv < 0.05)

# Sort pathways and set colors
snp_sort_ns <- snp_nsgenes %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(n = median(wmis_cv)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()
snp_sort_ns$color <- c("#8c2421", "#4f4fad", "#ca6627", "#6a9b85")

# Violin plot for NS genes with global mutations
snp_plot_nsmut <- ggviolin(snp_nsgenes, x = "pathway", y = "wmis_cv", color = "pathway",
                           shape = "pathway", xlab = "Pathway", ylab = "dN/dS",
                           order = snp_sort_ns$pathway, add = "jitter", legend = "none",
                           palette = snp_sort_ns$color) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  scale_y_continuous(limits = c(-0.2, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2)
ggsave("5-SNPs/dndscv/results/plot/plot_dnds_nsmut.pdf", snp_plot_nsmut, height = 9, width = 7)

# Merge PS and NS gene names
snp_gmerge <- bind_rows(snp_psgenes, snp_nsgenes)
snp_gmerge <- snp_gmerge %>%
  dplyr::mutate(pathway = ifelse(is.na(pathway), "PS", pathway))
#snp_color_gm <- c("#ddad3b", "#4f4fad", "#6a9b85", "#ca6627", "#8c2421", "black")
snp_color_gm <- c("#8c2421", "#4f4fad", "#6a9b85", "#ca6627")

# Lollipop plot dN/dS x-axis range 0-2, y-axis PS genes vs selected CDS
snp_plot_gmerge <- ggdotchart(snp_nsgenes, x = "gene_name", y = "wmis_cv", color = "pathway",
                              xlab = "Gene CDS", ylab = "dN/dS", y.text.col = FALSE, rotate = TRUE,
                              palette = snp_color_gm, sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 0.4)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
ggsave("5-SNPs/dndscv/results/plot/plot_dnds_nspath.pdf", snp_plot_gmerge, height = 9, width = 7)

# Genes for enrichment (we need KEGG KOs..)
#snp_genenrich <- snp_psgenes$dndsr
#names(snp_genenrich) <- snp_psgenes$name
#snp_kegg <- enrichKEGG(snp_genenrich, "hsa", pvalueCutoff = 0.05)

```

>>>>

```{r setup}
knitr::opts_knit$set(root.dir = "/Users/moises/Downloads/")

@@

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
#ggsave("fig_1.png", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b)

#fig sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
fig_s1[[2]][[1]] <- fig_s1[[2]][[1]] + theme(axis.title.y = element_text(vjust = 0))
#fig_s1[[2]][[2]] <- fig_s1[[2]][[2]] + theme(axis.text.y = element_text(vjust = 0))
ggsave("fig_s1.pdf", fig_s1, height = 8, width = 16)
#ggsave("fig_s1.png", fig_s1, height = 8, width = 16)
```

@@

```{r fig2}
# Merge plots using patchwork
fig_2 <- { { anova_boxplot_cu / anova_boxplot_fe +
    plot_layout(ncol = 1, widths = c(0.75)) } | anova_boxplot_s / kegg_bubbles_plot } +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d"))) &
  theme(plot.tag =  element_text(face = "bold"))
#adjust y-label margin for sulfur plot
fig_2[[2]][[1]] <- fig_2[[2]][[1]] + theme(axis.title.y = element_text(vjust = -127))
ggsave("fig_2.pdf", fig_2, height = 14, width = 18)
#ggsave("fig_2.png", fig_2, height = 14, width = 18) vjust=120
```

## Section 3: SNP Variant Calling

@@

```{r snps}
# Get CDS annotations (pre-processed from Bakta results)
snp_refcds <- read.csv("mags-all/5-SNPs/dndscv/buildref/cds_table.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Buildref testing for bacteria.. (1 bin)
#https://github.com/im3sanger/dndscv/issues/62
#buildref(cdsfile = "mags-all/5-SNPs/dndscv/buildref/cds_node_25854.txt",
#         genomefile = "mags-all/5-SNPs/dndscv/buildref/node_25854.fa",
#         outfile = "mags-all/5-SNPs/dndscv/buildref/testref.rda")
#load("mags-all/5-SNPs/dndscv/buildref/testref.rda")

# Buildref for all bins (very slow: ~2 hrs in building object)
#buildref(cdsfile = "mags-all/5-SNPs/dndscv/buildref/cds_table.txt",
#         genomefile = "mags-all/5-SNPs/dndscv/buildref/S15_bins.fa",
#         outfile = "mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda")

# Inspect refdb
#load("mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda")
#print(RefCDS[[1]])
#rm(gr_genes, RefCDS)

# Get annotations (parsed from the VCF generated with SnpEff)
#sampleID chr pos ref mut
snp_mutations <- read.csv("mags-all/5-SNPs/dndscv/run/S15_mutations.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Run dndscv (runtime ~8 min, high memory usage) and save object
#snp_dndsout <- dndscv(snp_mutations, refdb = "mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda",
#                      max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf,
#                      cv = NULL, outmats = TRUE)
#save(snp_dndsout, file = "mags-all/5-SNPs/dndscv/run/S15_dndsout.rda")
#save(substmodel, file = "mags-all/5-SNPs/dndscv/run/S15_substmodel.rda")
#load("mags-all/5-SNPs/dndscv/run/S15_dndsout.rda")
#load("mags-all/5-SNPs/dndscv/run/S15_substmodel.rda")
rm(snp_mutations)

# Explore output object..
#https://github.com/im3sanger/dndscv/issues/6
#https://github.com/im3sanger/dndscv/issues/64
head(snp_dndsout$sel_cv) #dN/dS ratios at gene-level
head(snp_dndsout$annotmuts)
print(snp_dndsout$globaldnds) #driver globals

# Save dN/dS per-gene matrix
#driver discovery or genes under positive selection (PS)
snp_selcv <- snp_dndsout$sel_cv
rm(snp_dndsout)

# Group global dN/dS ratios:
#https://github.com/im3sanger/dndscv/issues/63
snp_selcv %>% dplyr::filter(pmis_cv < 0.05) %>% nrow() #n=105328
snp_global_ps <- snp_selcv %>%
  dplyr::filter(wmis_cv > 1) %>%
  dplyr::summarise(genes = n())
snp_global_ps$label = "dN/dS > 1"
snp_global_ns <- snp_selcv %>%
  dplyr::filter(wmis_cv < 1 & wmis_cv != 0) %>%
  dplyr::summarise(genes = n())
snp_global_ns$label = "dN/dS < 1"
snp_global_dnds <- rbind(snp_global_ps, snp_global_ns)
snp_global_dnds <- snp_global_dnds %>%
  dplyr::mutate(perc = round(genes * 100 / nrow(snp_selcv), 2)) %>% #add %
  dplyr::mutate(perc2 = paste0(genes, " (", perc, "%)")) #label to plot
rm(snp_global_ps, snp_global_ns) #clean memory

# Filter genes under PS:
#https://github.com/im3sanger/dndscv/issues/56
#wmis_cv: dN/dS for missense mutations
#pmis_cv: p-value (signif. < 0.05) is expected to occur by chance under neutrality (null hypothesis)
snp_psgenes <- snp_selcv %>%
  dplyr::filter(wmis_cv > 1) %>%
  dplyr::mutate(dndsr = round((n_mis / n_syn), 1))

# Load contig to bin matrix
snp_bins <- read.csv("mags-all/5-SNPs/dndscv/results/S15_DASTool_contig2bin.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Load gene names (parsed from genes.gff file)
snp_genids <- read.csv("mags-all/5-SNPs/dndscv/results/gene_id_name.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

## FUNCTIONS ##
# Function that returns the contig where a gene was found
fx_dndsGene2Contig <- function(g) {
  c <- snp_refcds %>%
    dplyr::filter(gene.name == g) %>%
    dplyr::select(chr) %>%
    as.character()
  return(c)
}
# Function that returns the bin given a contig
fx_dndsContig2Bin <- function(c) {
  b <- snp_bins %>%
    dplyr::filter(contig == c) %>%
    dplyr::select(bin) %>%
    as.character()
  return(b)
}
# Function that gets gene names given an id
fx_dndsGeneName <- function(g) {
  g <- snp_genids %>%
    dplyr::filter(gene.id == g) %>%
    dplyr::select(gene.name) %>%
    as.character()
  return(g)
}
# Function that returns a row with the taxonomic classification of a bin
fx_dndsTaxaBin <- function(b) {
  t <- mag_samples %>%
    dplyr::mutate(bin = paste0(mag_name, ".", mag_id)) %>%
    dplyr::filter(bin == b) %>%
    dplyr::select(phylum, genus, species, predicted_genes)
  return(t)
}
# Function that process the product tag of NS genes
fx_dndsNsPathway <- function(p) {
  if (grepl("copper", p, ignore.case = TRUE) == TRUE)
    return("Copper")
  if (grepl("iron", p, ignore.case = TRUE) == TRUE)
    return("Iron")
  if (grepl("nickel", p, ignore.case = TRUE) == TRUE)
    return("Nickel")
  if (grepl("sulfur", p, ignore.case = TRUE) == TRUE)
    return("Sulfur")
  if (grepl("nitrogen", p, ignore.case = TRUE) == TRUE)
    return("Nitrogen")
}
# Function that gets dndscv stats for NS genes
fx_dndsNsStats <- function(g) {
  s <- snp_selcv %>%
    dplyr::filter(gene_name == g) %>%
    dplyr::select(gene_name, wmis_cv, pmis_cv)
  return(s)
}
## END FUNCTIONS ##

@@

# Group PS genes to add taxa & QC
snp_dnds <- snp_psgenes %>%
  dplyr::group_by(bin) %>%
  dplyr::summarise(num = n()) %>%
  dplyr::arrange(desc(num))

snp_taxabins <- tibble(phylum = character(), genus = character(), species = character(),
                       completeness = numeric(), contamination = numeric())
for (i in 1:nrow(snp_dnds)) {
  #print(sprintf("%i -> bin: %s", i, fx_dndsTaxaBin(snp_dnds$bin[i])))
  snp_taxabins <- rbind(snp_taxabins, fx_dndsTaxaBin(snp_dnds$bin[i]))
}
snp_dnds['phylum'] <- snp_taxabins$phylum
snp_dnds['genus'] <- snp_taxabins$genus
snp_dnds['species'] <- snp_taxabins$species
snp_dnds['predicted_genes'] <- snp_taxabins$predicted_genes
rm(snp_taxabins)

@@

# Subset PS genes to plot mutations -> synonymous/missense (non-syn.)/nonsense
#discard HPs (by name)
#genes with at least 5 SNPs
#mark genes reaching significance (q < 0.05)
#mark genes: KJCAEPPB_04437 -> nickel pathway; NEBJFEFK_01910 -> sulfur pathway
snp_psgenes_sub <- snp_psgenes %>%
  dplyr::filter(nchar(name) < 14) %>%
  dplyr::mutate(snps = n_syn + n_mis + n_non) %>%
  dplyr::filter(snps >= 4) %>%
  dplyr::mutate(signif = ifelse(pmis_cv < 0.05, "*", "")) %>%
  dplyr::mutate(path = case_when(gene_name == "KJCAEPPB_04437" ~ "Ni",
                                 gene_name == "NEBJFEFK_01910" ~ "S", TRUE ~ "")) %>%
  distinct(name, .keep_all = TRUE) %>% #remove repeated names
  pivot_longer(cols = c(n_syn, n_mis, n_non), names_to = "mut", values_to = "n_mut")

# Save PS gene names and ids to file
if (file.exists("mags-all/5-SNPs/dndscv/results/ps_genes.txt") == FALSE) {
  write.table(snp_psgenes_sub %>% dplyr::select(gene_name) %>% distinct(gene_name), "mags-all/5-SNPs/dndscv/results/ps_genes.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}
if (file.exists("mags-all/5-SNPs/dndscv/results/kofam/ps_genes_ids.txt") == FALSE) {
  write.table(snp_psgenes_sub %>% dplyr::select(name) %>% distinct(name), "mags-all/5-SNPs/dndscv/results/kofam/ps_genes_ids.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

# Barplot with global dN/dS ratios
snp_plot_dndsglob <- ggplot(snp_global_dnds, aes(x = genes, y = reorder(label, genes), fill = label)) +
  geom_col(position = "dodge2", fill = "#4c6194") +
  geom_text(aes(label = perc2), vjust = 0.5, hjust = -0.1, size = 2) +
  scale_x_continuous(limits = c(0, 16000)) +
  labs(x = "Number of genes", y = "") +
  theme_classic() +
  theme(legend.position = "none")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_global.pdf", snp_plot_dndsglob, height = 2, width = 5)

# Plot global driver genes by species
snp_plot_psbins <- ggplot(snp_dnds, aes(x = num, y = reorder(bin, num), fill = bin)) +
  geom_bar(stat = "identity", fill = "#E31A1C") +
  geom_text(aes(label = num), vjust = 0.5, hjust = -0.5, size = 2.5) +
  labs(x = "Number of genes under PS", y = "Bin") +
  scale_x_continuous(limits = c(0, 255)) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  facet_grid(.~species)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 8.5, width = 9.5)

# Lollipop plot of PS genes
snp_plot_psgenes <- ggplot(snp_psgenes_sub,
                           aes(x = wmis_cv, y = reorder(name, wmis_cv)), size = wmis_cv) +
  geom_segment(aes(x = 0, xend = wmis_cv, y = name), color = "gray") +
  geom_point(alpha = 0.7, color = "#E31A1C") +
  geom_text(aes(label = signif, x = -0.15), size = 3, color = "#3c4d76") +
  geom_text(aes(label = path, x = -0.15), size = 2, color = "#3c4d76") +
  geom_vline(xintercept = 1, linetype = 1, linewidth = 0.2) +
  geom_vline(xintercept = max(snp_psgenes_sub$wmis_cv), linetype = 1, linewidth = 0.2) +
  labs(x = "PS strength", y = "Gene CDS") +
  theme_classic()
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psgenes.pdf", snp_plot_psgenes, height = 11, width = 7)

# Plot PS genes mutations per type
snp_plot_muts <- ggplot(snp_psgenes_sub, aes(x = n_mut, y = reorder(name, wmis_cv), fill = mut)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("mis", "non", "syn"),
                    values = c("#5db4b2", "#b687b6", "#2374a4")) +
  labs(x = "Number of mutations", y = "") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.15),
        strip.placement = "outside",
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psmuts.pdf", snp_plot_muts, height = 11, width = 7)

snp_plot_psgenes <- (snp_plot_psgenes | snp_plot_muts) +
  plot_layout(nrow = 1, ncol = 2, widths = c(0.55, 0.5))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psgenmuts.pdf", snp_plot_psgenes, height = 11, width = 11)
rm(snp_plot_muts)

# Load negative selected (NS) genes
snp_nsgenes <- read.csv("mags-all/5-SNPs/dndscv/results/selected_cds_metals.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Infer pathway from product and add dnds attributes
snp_nspath <- tibble(gene_name = character(), pathway = character())
snp_nsstat <- tibble(wmis_cv = numeric(), pmis_cv = numeric())
for (i in 1:nrow(snp_nsgenes)) {
  #use dataframes, not lists
  snp_nspath <- snp_nspath %>%
    add_row(gene_name = snp_nsgenes$gene_name[i], pathway = fx_dndsNsPathway(snp_nsgenes$product[i]))
	snp_nsstat <- rbind(snp_nsstat, fx_dndsNsStats(snp_nsgenes$gene_name[i]))
}
snp_nsgenes <- left_join(snp_nsgenes, snp_nspath, by = c("gene_name"))
snp_nsgenes <- left_join(snp_nsgenes, snp_nsstat, by = c("gene_name"))
rm(snp_nspath, snp_nsstat)

# Filter NS genes not called in dndscv
#omit significance (pmis_cv < 0.05)
#discard nitrogen (not enough genes)
#change %2C to comma
snp_nsgenes <- snp_nsgenes %>%
  dplyr::filter(!(is.na(wmis_cv)) & wmis_cv < 1 & wmis_cv != 0 & pathway != "Nitrogen") %>%
  dplyr::mutate(product = str_replace_all(product, "%2C", ","))

# Save NS gene ids to file
if (file.exists("mags-all/5-SNPs/dndscv/results/kofam/ns_genes_ids.txt") == FALSE) {
  write.table(snp_nsgenes$gene_id, "mags-all/5-SNPs/dndscv/results/kofam/ns_genes_ids.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
}

# Sort pathways and set colors to plot
snp_sort_ns <- snp_nsgenes %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(n = mean(wmis_cv)) %>%
  dplyr::arrange(desc(n)) %>%
  as.data.frame()
snp_sort_ns$color <- c("#ddad3b", "#8c2421", "#666666", "#4f4fad") #"#6a9b85"

# Violin plot for NS genes with global mutations
snp_plot_nsgenes <- ggviolin(snp_nsgenes, x = "pathway", y = "wmis_cv", color = "pathway",
                             shape = "pathway", xlab = "Pathway", ylab = "dN/dS",
                             order = snp_sort_ns$pathway, add = "jitter", legend = "none",
                             palette = snp_sort_ns$color) +
  scale_y_continuous(limits = c(-0.2, 1.2)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  stat_compare_means(method = "anova", label.y = 1.2) +
  stat_compare_means(comparisons = list(c("Copper", "Iron"), c("Sulfur", "Nickel")),
                     label = "p.format", method = "t.test")
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenes.pdf", snp_plot_nsgenes, height = 9, width = 7)

# Merge NS genes to unique products
snp_nsgenes_sub <- snp_nsgenes %>%
  dplyr::group_by(product, pathway) %>%
  dplyr::summarise(dnds = mean(wmis_cv))
snp_nsgenes_sub_col <- c("#8c2421", "#666666", "#4f4fad", "#ddad3b")

# Lollipop plot of NS genes dN/dS x-axis range 0-2, y-axis PS genes vs selected CDS
snp_plot_nsgensub <- ggdotchart(snp_nsgenes_sub, x = "product", y = "dnds", color = "pathway",
                              xlab = "NS gene product", ylab = "dN/dS", y.text.col = FALSE,
                              rotate = TRUE, palette = snp_nsgenes_sub_col,
                              sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenespath.pdf", snp_plot_nsgensub, height = 9, width = 10)
